|

### 2023-01-06 review/response.Rmd

---
title: "Response to review: New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions"
author: "Ursula Laa, Alex Aumann, Dianne Cook, German Valencia"
date: "2023-01-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Editor**

A referee, an Associate Editor (AE), and I have reviewed your paper. There's enthusiasm for your paper but a number of issues that came up in the reviews require additional attention.  As such, I'd like to offer the opportunity to submit a revised version that addresses the issues brought up by the AE and referee.

1- Your revision should not increase the length of the manuscript.

2- I will download and run your code as part of the next round of reviews.  I suggest, but don't require, the tidyverse style for your code and, in any case, ask that you submit code that is documented and as user-friendly as possible.  Before resubmitting please run your code in a "clean" environment as issues such as required packages that aren't loaded or calls to files that don't exist are common and can delay publication (I'm not saying these exist in your case, just asking you to be aware!).

3- Your paper will be printed in black and white.  The online version will be in color but to accommodate printing without color please be sure you don't refer to color in either the figure captions or text and that all figures are still interpretable in black and white.

4- Please be sure all figures have font size of labels and any text in the figure of at least 10pt.

Articles published in JCGS are ordinarily accompanied by online publication of the data and computer code used to illustrate the proposed methods. The goal is to enable readers to replicate much of the analysis and simulations described in a paper and to apply proposed methods in their own data without having to write code from scratch.  When it is not practical to include a whole dataset in the manuscript, the manuscript should state how the complete dataset can be obtained. Please let me know if you would not be able to publish your data for reasons of security or confidentiality. Please provide the computer code and/or data that you hope to publish along with your revision. I have attached the instructions for including online materials here.

**Associate Editor**

Summary
This paper describes a method for interacting with multidimensional data using slice tours in Mathematica (as well as a reduced-feature implementation in R).

Major Issues
- Reproducibility:
Code should be reproducible and not depend on user paths.
Data files should be uploaded along with code files (they don’t seem to be part of supplemental material?)
Not having the ability to actually run the code (because the data wasn’t included) makes it somewhat diﬀicult to assess the software side of this paper.
- Grammar: There are a lot of minor grammatical issues with this paper. I suggest utilizing a tool such as Grammarly to spot and fix them. The ones that I noticed as I read through the first section are detailed in minor issues below, but I am trying to focus more on the content than the style, so I will try to not nit-pick through the rest of the paper and trust that the authors can use the online tools.
- Cite your software packages appropriately - they're tools you depend on to do your work and should be cited in the same way papers are cited.
– Page 3, line 28: Capitalize Mathematica and cite it appropriately. – Page 3, line 32: Cite R
– Page 7, line 28: Cite Shiny
- Notation
– Define your notation better: $V_m$ is used without being defined, and it would proba- bly be helpful to have a clearer mathematical definition of the relationship between your basis $A$ and $V_m$.

- What developments are specific to Mathematica and are not yet implemented in R? You allude several times to R’s deficiencies in this space but don’t ever explicitly spell out what is missing, beyond a nod to mouse tracking capabilities in Shiny (which would likely depend on the user’s screen resolution, but should be fairly precise once you condition on that - why would Mathematica have better resolution for mouse controls?)
Below is a set of comments relating to this issue, written as I read through the paper (so hopefully the chronology is preserved and you can see why this isn’t as direct as it
should be).

Page 7, line 46: "Mathematica provides much utility and versatility, such as inbuilt data visualization, data manipulation and analysis, dynamic functionality, and sym- bolic and numeric computation. Most of this inbuilt functionality is user-friendly and described in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there."
The same could be said of R, so I’m not sure why this is relevant. It seems mostly like Mathematica propaganda.
– Page 7-8: "Importantly for our work, it is relatively simple to create dynamic objects via the inbuilt commands Manipulate or DynamicModule. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated. This is the essential ingredient for this implementation of the manual tour."
Again, the same dynamic control creation exists in Shiny, along with the basic reactivity to user inputs. It's hard to see why Mathematica is different here.
– Page 8, line 11: "The most relevant inbuilt Mathematica function for our purpose is the LocatorPane. This creates a region on the screen where the position of the mouse is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour."
Finally, after all of this time, this is the one input that might be a bit harder to implement in R/Shiny, though I think via WebGL it should be workable. This is where the focus should be, rather than the previous two paragraphs, which are a distraction.

Minor issues and suggestions

- Pg 2, line 8: "the more useful methods for data analysis need to work for arbitrary dimension."
Perhaps instead: methods which work for arbitrary dimensions are more useful/have more general utility? Either way, dimension should be plural.
- Pg 2, line 17: "could theoretically show the viewer the data from all sides."
It might be helpful to have a small figure here to demonstrate what this might mean, for those who aren't familiar with tours and/or who can't visualize n-dimensional space projections into 3D out of thin air. Alternately, link to e.g. a YouTube video showing what you mean by "from all sides"
- Pg 2, line 23: "A video illustrating the capabilities is available through video library of ASA Statistical Graphics Section (2022)."
Add "the" before video

**Reviewer 1**

In this paper a new manual controls for projections and slice tours are presented. The controls are designed to explore high-dimensional data based on linear projections and slices.  In the paper a Mathematica package is described to explain the application.

Some comments:

In the Abstract you have mentioned “Some limited implementations has also been made available in the R package tourr” but in Sub-Section 3.2 you can mention in more detail the limitations in tourr implementation

Section 1 line 29 “The software section describes mathematica package”  mathematica should be capitalized as you did  through the paper in most cases.  Also since is the first time you have mentioned Mathematica you should include the reference here (Wolfram reference appears in page 4 line 26 for first time)

Section 2: “manual tour allows the user to alter the coefficients of one (or more) variables” but in the explanation it is not clear what is  V_m  and how to alter the coefficients for more than one variable.
In page 4 line 2.2 “The components corresponding to V_m are directly controlled by cursor movement which updates row m of A ''. Maybe a video will be illustrative in this sub-section.

Section 3: will be useful to include a schematic diagram illustrating the interactivity 
In the manual tour. implementation .
This part is difficult to follow, it is more clear after reading the application Section because you have included some videos but I think you need to include  something here to help the reader.

In the introduction “The software section describes a mathematica package that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available ”
I think you need to be more specific about the R limitations.
Iin Section 3. Page 7, line 20 “One might implement such an interactive interface via R Shiny, but in particular the tracking of the mouse position in small increments might pose a challenge” This  is the only possible R limitation you have mentioned.

Section 4: Page 10 line 36: Figure 3summarizes (space)

Section 5: page 15 line 26 “If new technology for interactive graphics became available”  be more specific about the needed technology in R to implement the manual controls as you did with Mathematica

### 2023-01-06 review/response.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Editor**

A referee, an Associate Editor (AE), and I have reviewed your paper. There's enthusiasm for your paper but a number of issues that came up in the reviews require additional attention.  As such, I'd like to offer the opportunity to submit a revised version that addresses the issues brought up by the AE and referee.

\textcolor{blue}{Thank you for the careful review of our paper. We have done our best to make changes to address each comment, as detailed in blue below.}

@@

4- Please be sure all figures have font size of labels and any text in the figure of at least 10pt.

Articles published in JCGS are ordinarily accompanied by online publication of the data and computer code used to illustrate the proposed methods. The goal is to enable readers to replicate much of the analysis and simulations described in a paper and to apply proposed methods in their own data without having to write code from scratch.  When it is not practical to include a whole dataset in the manuscript, the manuscript should state how the complete dataset can be obtained. Please let me know if you would not be able to publish your data for reasons of security or confidentiality. Please provide the computer code and/or data that you hope to publish along with your revision. I have attached the instructions for including online materials here.

**Associate Editor**

Summary

This paper describes a method for interacting with multidimensional data using slice tours in Mathematica (as well as a reduced-feature implementation in R).

Major Issues

- Reproducibility:
    - Code should be reproducible and not depend on user paths.
    - Data files should be uploaded along with code files (they don't seem to be part of supplemental material?)
    - Not having the ability to actually run the code (because the data wasn't included) makes it somewhat diﬀicult to assess the software side of this paper.
- Grammar: There are a lot of minor grammatical issues with this paper. I suggest utilizing a tool such as Grammarly to spot and fix them. The ones that I noticed as I read through the first section are detailed in minor issues below, but I am trying to focus more on the content than the style, so I will try to not nit-pick through the rest of the paper and trust that the authors can use the online tools.
- Cite your software packages appropriately - they're tools you depend on to do your work and should be cited in the same way papers are cited.

    – Page 3, line 28: Capitalize Mathematica and cite it appropriately.

    – Page 3, line 32: Cite R

    – Page 7, line 28: Cite Shiny

- Notation

    – Define your notation better: $V_m$ is used without being defined, and it would probably be helpful to have a clearer mathematical definition of the relationship between your basis $A$ and $V_m$.

- What developments are specific to Mathematica and are not yet implemented in R? You allude several times to R's deficiencies in this space but don't ever explicitly spell out what is missing, beyond a nod to mouse tracking capabilities in Shiny (which would likely depend on the user's screen resolution, but should be fairly precise once you condition on that - why would Mathematica have better resolution for mouse controls?)
Below is a set of comments relating to this issue, written as I read through the paper (so hopefully the chronology is preserved and you can see why this isn’t as direct as it
should be).

    - Page 7, line 46: "Mathematica provides much utility and versatility, such as inbuilt data visualization, data manipulation and analysis, dynamic functionality, and sym- bolic and numeric computation. Most of this inbuilt functionality is user-friendly and described in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there." The same could be said of R, so I'm not sure why this is relevant. It seems mostly like Mathematica propaganda.

        – Page 7-8: "Importantly for our work, it is relatively simple to create dynamic objects via the inbuilt commands Manipulate or DynamicModule. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated. This is the essential ingredient for this implementation of the manual tour." Again, the same dynamic control creation exists in Shiny, along with the basic reactivity to user inputs. It's hard to see why Mathematica is different here.

        – Page 8, line 11: "The most relevant inbuilt Mathematica function for our purpose is the LocatorPane. This creates a region on the screen where the position of the mouse is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour."

    - Finally, after all of this time, this is the one input that might be a bit harder to implement in R/Shiny, though I think via WebGL it should be workable. This is where the focus should be, rather than the previous two paragraphs, which are a distraction.

@@

**Reviewer 1**

In this paper a new manual controls for projections and slice tours are presented. The controls are designed to explore high-dimensional data based on linear projections and slices.  In the paper a Mathematica package is described to explain the application.

Some comments:

In the Abstract you have mentioned "Some limited implementations has also been made available in the R package tourr" but in Sub-Section 3.2 you can mention in more detail the limitations in tourr implementation

Section 1 line 29 "The software section describes mathematica package"  mathematica should be capitalized as you did  through the paper in most cases.  Also since is the first time you have mentioned Mathematica you should include the reference here (Wolfram reference appears in page 4 line 26 for first time)

Section 2: "manual tour allows the user to alter the coefficients of one (or more) variables" but in the explanation it is not clear what is  $V_m$  and how to alter the coefficients for more than one variable.

In page 4 line 2.2 "The components corresponding to $V_m$ are directly controlled by cursor movement which updates row $m$ of $A$". Maybe a video will be illustrative in this sub-section.

Section 3: will be useful to include a schematic diagram illustrating the interactivity

In the manual tour implementation.
This part is difficult to follow, it is more clear after reading the application Section because you have included some videos but I think you need to include something here to help the reader.

In the introduction "The software section describes a mathematica package that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available"
I think you need to be more specific about the R limitations.
Iin Section 3. Page 7, line 20 "One might implement such an interactive interface via R Shiny, but in particular the tracking of the mouse position in small increments might pose a challenge" This  is the only possible R limitation you have mentioned.

Section 4: Page 10 line 36: Figure 3 summarizes (space)

Section 5: page 15 line 26 "If new technology for interactive graphics became available"  be more specific about the needed technology in R to implement the manual controls as you did with Mathematica

### 2023-01-11 review/response.Rmd

    – Page 3, line 32: Cite R

    – Page 7, line 28: Cite Shiny
\textcolor{blue}{done}    
- Notation

    – Define your notation better: $V_m$ is used without being defined, and it would probably be helpful to have a clearer mathematical definition of the relationship between your basis $A$ and $V_m$.

\textcolor{blue}{see text in blue page 3, is this adequate?}

@@

      – Page 7-8: "Importantly for our work, it is relatively simple to create dynamic objects via the inbuilt commands Manipulate or DynamicModule. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated. This is the essential ingredient for this implementation of the manual tour." Again, the same dynamic control creation exists in Shiny, along with the basic reactivity to user inputs. It's hard to see why Mathematica is different here.

        – Page 8, line 11: "The most relevant inbuilt Mathematica function for our purpose is the LocatorPane. This creates a region on the screen where the position of the mouse is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour."

    - Finally, after all of this time, this is the one input that might be a bit harder to implement in R/Shiny, though I think via WebGL it should be workable. This is where the focus should be, rather than the previous two paragraphs, which are a distraction.

\textcolor{blue}{The first two paragraphs have been removed and the third restructured to explain why this is difficult in R/Shiny}

@@

Minor issues and suggestions

- Pg 2, line 8: "the more useful methods for data analysis need to work for arbitrary dimension."
Perhaps instead: methods which work for arbitrary dimensions are more useful/have more general utility? Either way, dimension should be plural.

\textcolor{blue}{done}

- Pg 2, line 17: "could theoretically show the viewer the data from all sides."
It might be helpful to have a small figure here to demonstrate what this might mean, for those who aren't familiar with tours and/or who can't visualize n-dimensional space projections into 3D out of thin air. Alternately, link to e.g. a YouTube video showing what you mean by "from all sides"

- Pg 2, line 23: "A video illustrating the capabilities is available through video library of ASA Statistical Graphics Section (2022)."
Add "the" before video
\textcolor{blue}{done}

@@

In the Abstract you have mentioned "Some limited implementations has also been made available in the R package tourr" but in Sub-Section 3.2 you can mention in more detail the limitations in tourr implementation

Section 1 line 29 "The software section describes mathematica package"  mathematica should be capitalized as you did  through the paper in most cases.  Also since is the first time you have mentioned Mathematica you should include the reference here (Wolfram reference appears in page 4 line 26 for first time) 
\textcolor{blue}{done}

Section 2: "manual tour allows the user to alter the coefficients of one (or more) variables" but in the explanation it is not clear what is  $V_m$  and how to alter the coefficients for more than one variable.

\textcolor{blue}{definition of $V_m$ added but have not addressed the second part of the comment}

### 2023-01-11 paper/paper.Rmd
start fixing errors

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods is illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a mathematica package that is used for the application, and describes the interactive environment that would be desirable within R as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and three Mathematica notebooks that can be used to reproduce the application.

@@

A manual tour allows the user to alter the coefficients of one (or more) variables contributing to a $d$ dimensional projection. The initial ingredients are an orthonormal basis ($A_{p\times d}$) defining the projection of the data, and a variable id ($m \in \{1, ..., p\}$) specifying which coefficient will be changed. A method to update the values of the component ($m^{th}$ row of $A_{p\times d}$) of the controlled variable $V_m$ is then needed.

@@

One might implement such an interactive interface via R Shiny, but in particular the tracking of the mouse position in small increments might pose a challenge. Here we found the dynamic graphics interface from Mathematica to be especially useful. In the following we first describe the relevant Mathematica functionalities and how they were used in the implementation, followed by a brief sketch of how the new approaches are implemented in the `tourr` package. There is hope that better interactivity will be available for R soon, which might allow for implementation of the techniques developed in Mathematica.

@@

Mathematica provides much utility and versatility, such as inbuilt data visualization, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is user-friendly and described in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there.

Importantly for our work, it is relatively simple  to create dynamic objects via the inbuilt commands `Manipulate` or `DynamicModule`. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated. This is the essential ingredient for this implementation of the manual tour.

The most relevant inbuilt Mathematica function for our purpose is the `LocatorPane`. This creates a region on the screen where the position of the mouse is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour.

>>>>

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, \textcolor{blue}{methods which work for arbitrary dimensions have more general utility}. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through \textcolor{blue}{the} video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

This paper explains the new manual controls for projection and slice tours. The next section describes the new algorithm for manual control, for both projections and slices. The use of these methods is illustrated to compare and contrast boundaries constructed by different classifiers. The software section describes a Mathematica @Mathematica package that is used for the application, and describes the interactive environment that would be desirable within R @rref as new technology becomes available. The paper is accompanied by an appendix with more details and adjustments to the manual controls, and three Mathematica notebooks that can be used to reproduce the application.

@@

A manual tour allows the user to alter the coefficients of one (or more) variables contributing to a $d$ dimensional projection. The initial ingredients are an orthonormal basis ($A_{p\times d}$) defining the projection of the data, and a variable id ($m \in \{1, ..., p\}$) specifying which coefficient will be \textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.} A method to update the values of the component of the controlled variable $V_m$ is then needed.

@@

One might implement such an interactive interface via R Shiny @shinym, but in particular the tracking of the mouse position in small increments might pose a challenge. Here we found the dynamic graphics interface from Mathematica to be especially useful. In the following we first describe the relevant Mathematica functionalities and how they were used in the implementation, followed by a brief sketch of how the new approaches are implemented in the `tourr` package. There is hope that better interactivity will be available for R soon, which might allow for implementation of the techniques developed in Mathematica.

@@

\textcolor{blue}{The Mathematica implementation relies on the inbuilt Mathematica function  `LocatorPane`. This function creates a region on the screen where the position of the mouse is captured and then converted to input that updates the graphics functions, enabling the manual navigation of the tour. We then use the graphics function `ListPlot` embedded in dynamic objects via  `DynamicModule` to produce the display. Control objects, such as sliders, locator panes, and input fields, are used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated.

Di will explain what part and why is difficult in R, also check grammar}
